Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-14 writeup/conStruct.tex
.

\begin{abstract}
A classic problem in population genetics is the characterization 
of discrete population structure in the presence of 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
clustering or assignment methods may incorrectly ascribe differentiation 
due to continuous processes (e.g., isolation by distance) 
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
This is partly a result of the difficulty of sampling uniformly and continuously 
across the range of a population or species, 
but more, it reflects a shortcoming of current methods for inferring and 
visualizing population structure from genetic data in the face of data 
that are characterized by both continuous and discrete population structure.
Here, we present a novel statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of discrete population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This model explicitly addresses the ``clines vs. clusters" problem in 
modeling population genetic variation by jointly accommodating both 
continuous and discrete patterns of differentiation. 
We demonstrate the utility of this approach using both simulations and empirical applications.
\end{abstract}

\newpage
\section*{Introduction}
A fundamental quandary in the description of biological diversity is
the fact that diversity shows both discrete and continuous patterns. 
For example, reasonable people can disagree about whether two 
populations are separate species, because the 
process of speciation is usually gradual, and so there is no set point
in the continuous divergence of populations when they unambiguously become discrete species.
The issue of identifying meaningful biological subunits 
extends below the species level, as patterns of phenotypic
and genetic diversity within and among populations are shaped
by continuous migration and drift, as well as by more discrete events, 
such as geographic barriers, rapid expansions, bottlenecks, and rare
long-distance migration.

From a practical standpoint, even while acknowledging continuous
processes, we often need to identify somewhat separable populations 
from which individuals are sampled \citep{wright1949genetical}.
Delineating populations is useful for systematics and for
informing conservation priorities \citep{Moritz1994,Waples_1998,Moritz_etal_2002}.
Furthermore, we often need to identify subsets of indivduals resulting from reasonably coherent
evolutionary histories for downstream analyses to learn about population history and adaptation.

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \citep{CavalliSforza1975, treemix},
dimensionality-reduction approaches \citep{meirmans2009genodive}, 
such as principal components analysis 
\citep{menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches 
\citep[e.g.][]{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, 
FINESTRUCTURE, fastStructure, huelsenbeck2007inference, 
Corander2003,TESS,geneland}, among others.
Each of these methods perform best in particular situations,
but many can give misleading results when applied to data 
that show a continuous pattern of isolation by distance 
\citep{Wright1943, novembre_interpreting_2008, Frantz2009}.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Existing model-based clustering methods model each individual's genotypes 
as random draws from a set of underlying, unobserved population clusters, 
each with a characteristic set of allele frequencies, which are estimated. 
These underlying frequencies are identical 
for all individuals assigned to a cluster, 
\emph{regardless} of their spatial location. 
Spatial information has been incorporated into some of these methods, 
by, for example, placing spatial priors on cluster memberships \citep{geneland,TESS}, 
but this does not address the underlying issue that 
these methods assume that allele frequencies 
are constant in a cluster across the species' range.

A pattern of isolation by distance (IBD) arises when migration in a
population is spatially limited, and slow enough between distant
locations that genetic drift can build up differentiation between them
\citep{Wright1943}. 
IBD has been modeled in both spatially continuous frameworks 
\citep{malecot1969mathematics,Slatkin1985}, 
in which an individual's parents are found within some radius of that individual's location,
and discrete, ``stepping-stone'' frameworks \citep{kimura1964stepping},
in which migration occurs only between neighboring demes.
Given the generality of the circumstances that generate a pattern of isolation by distance, 
it is unsurprising that IBD is very widespread in nature \citep{meirmans2012,Sexton_etal_2014}.

@@

To address this, we set out to develop
a model-based clustering method that, when possible, uses isolation by distance 
to explain observed genetic variation.
With an explicit spatial component, discrete population structure need only be invoked when genetic differentiation 
in the data deviates significantly from that expected given geographic separation.

In this paper, we present a statistical method for simultaneously 
modeling continuous and discrete patterns of population structure.
We model genetic variation in genotyped individuals as 
partitioned within or admixed across a specified number of discrete layers,
within each of which relatedness decays as a parametric function of the distance between samples.
We also implement a cross-validation approach for comparing and selecting models across different numbers of layers,
and we demonstrate the utility of our approach using both simulated and empirical data.
The implementation of this method, \texttt{conStruct} (for ``\emph{con}tinuous \emph{struct}ure"), 
is available as an R package at 
\href{https://github.com/gbradburd/conStruct}{https://github.com/gbradburd/conStruct}.

@@

\paragraph{Data}
The statistical framework of our approach is conceptually similar to that in \cite{Wasser2004}, \cite{BEDASSLE}, and \cite{spacemix},
although we use a somewhat different summary statistic than in this previous work.
The model works with allele frequencies at $L$ unlinked, bi-allelic single nucleotide polymorphisms (SNPs) genotyped across $N$ samples.
Each ``sample'' may be a single individual,  
a collection of individuals from a location, 
or frequencies estimated from pooled sequencing.
We write $N_i$ for the number of chromosomes sequenced in the $i^\text{th}$ sample.
The \emph{sample frequency} at locus $\ell$ in sample $n$, denoted $f_{n,\ell}$, 
is calculated by first arbitrarily choosing one of the observed alleles at locus $\ell$ to count, 
then dividing the number of observations of that counted allele by the total number of chromosomes genotyped at that locus
in sample $n$.
The choice of allele does not affect subsequent calculations, and so may be arbitrary.
We then calculate the \emph{allelic covariance} between samples $i$ and $j$, denoted $\widehat{\Omega}_{i,j}$,
as the expected covariance of distinct individual alleles chosen from each of the two samples at a random locus,
choosing a random allele to code as `1' (and the other as `0').
Concretely, this is
\begin{equation}
\widehat{\Omega}_{i,j} = 
    \frac{1}{L} \sum_{\ell=1}^L \left( (f_{i,\ell}-1/2) (f_{j,\ell}-1/2) - \delta_{i,j} f_{i,\ell}^2 / (N_i-1) \right),
\label{allelic_covariance}
\end{equation}
where $\delta_{i,j}=1$ if $i=j$ and is 0 otherwise.
Although we describe this as a covariance between individually drawn alleles,
$\widehat{\Omega}_{i,j}$ is in fact also the covariance between the allele frequencies
of a randomly chosen allele in samples $i$ and $j$, as long as $i \neq j$.
The diagonal (where $i=j$) does not have this interpretation, 
and reflects the covariance between individual alleles drawn from within the population.

@@

\paragraph{Continuous and discrete differentiation}
Clustering approaches to describing genetic variation
are useful because population history can often be meaningfully described on a coarse scale
by interactions between discrete ``populations''
whose relationships are delimited by patterns of glaciation,
large-scale migration,
mountain ranges, and the like.
Here we add a spatial component within each such discrete historical component,
which we refer to as a set of ``layers'' that overlay the modern map.
We imagine each layer as a geographically distributed population 
that extends over the entire sampled range of the populations. 
Each sample is composed of a mixture of contributions from each of
these layers. 
See Figure \ref{schematic} for a cartoon representation of our model. 
These layers thus take the place of ``clusters'' in clustering methods,
but we do not adopt this term, 
as ``spatial cluster'' suggests a clustering in space, 
while our layers may be present at all sampled locations.

@@

Within each of these layers,
allele frequencies have positive covariance at geographically close locations,
but this covariance is allowed to decay as geographic distance increases.
This pattern of spatial decay reflects how migration between neighboring demes  
homogenizes allele frequency changes that arise locally due to drift, 
but less effectively homogenizes geographic distant demes,
resulting in a continuous pattern of isolation by distance within each layer.
There is a fixed amount of covariance between layers, irrespective of spatial location.
Within each layer allele frequencies are expected to change gradually with distance,
but observed frequencies can change abruptly at many loci 
if the proportions by which individuals' ancestry derive from each layer 
(the ``admixture'' coefficients) 
do so as well.

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each layer, 
we define the covariance within layer $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_layer_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th layer.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the layer.
Our choice of a powered-exponential decay, 
as parameterized by the $\alpha$s, 
is a flexible and standard choice in spatial statistics \citep{Diggle1998}, 
and is not chosen to match a particular population genetics model. 
The $\phi^{(k)}$ is a parameter that describes the background covariance within the layer. 
If two samples draw 100\% of their ancestry from layer $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the layer.
The ``shared drift'' parameter $\phi^{(k)}$ can be interpreted as the branch length 
connecting the $k$th population to the population ancestral to all modeled
layers \citep[see for example ][]{patterson_ancient_2012,
  peter_fstats}. In the methods section \ref{rationale} we lay out a
simple model of allele frequencies underlying this covariance model.

We then allow samples to draw their ancestry from more than one layer.
The ``admixture'' proportion of the $i$th sample in the $k$th layer, denoted $w^{(k)}_i$,
gives the genome-wide proportion of alleles from sample $i$ that derive from
layer $k$ (and so $\sum_K w^{(k)}_i =1$).
A visual representation of the method mechanics is shown in Fig \ref{schematic}.

\begin{figure}
	\centering
		{\includegraphics[width=4in,height=2.5in]{figs/schematic/method_schematic.png}}
		\caption{Schematic of our method, using $K=3$ as an example.
			     The pattern of isolation by distance within each layer is denoted by their color gradients, 
			     and $\phi_{(k)}$ denotes the covariance shared by samples with ancestry entirely in the $k$th layer.
			     Sampled populations on the landscape are inferred to be admixed between these layers; 
			     the $i$th sample draws proportion $w^{(k)}_i$ of its ancestry from layer $k$.
			     For convenience, each layer is depicted as a small square, but in fact, 
			     each layer exists everywhere in the sampled area,
			     so the small dashed circles on each layer show where 
			     the location of the highlighted, admixed sample intersects each layer.
			    }\label{schematic}
\end{figure}

We can then describe the covariance between samples $i$ and $j$ across all layers, $\Omega_{i,j}$,
by summing their within-layer spatial covariances ($G_{i,j}^{(k)}$ in layer $k$) across all $K$ layers,
weighted by the relevant admixture proportions.
\begin{equation}
\Omega_{i,j} = \gamma + \sum\limits_K w^{(k)}_i w^{(k)}_j
G^{(k)}_{i,j} + \delta_{i,j}\eta_i .
\label{cross_layer_covariance}
\end{equation}
In this equation, $w^{(k)}_i w^{(k)}_j$ is the proportion of alleles that \emph{both}
sample $i$ and sample $j$ have inherited from layer $k$.

@@

\paragraph{Likelihood and inference}
Assuming that the allele frequencies at each locus
are independent between loci and multivariate normally distributed across populations, 
their allelic covariance $\widehat{\Omega}$ 
would be Wishart distributed with degrees of freedom equal to $L$, 
the number of loci genotyped.
We use this as a convenient approximation to the true distribution described above,
and so define the likelihood of the allelic covariance to be
\begin{equation}
P(\widehat{\Omega} \; | \; \Omega) = \mathcal{W} \left( L\widehat{\Omega} \; | \; \Omega,L\right) .
\end{equation}
Linkage disequilibrium (LD) between loci will decrease the effective
number of degrees of freedom.  
One possible solution, which we have not yet implemented, 
is to estimate an \emph{effective} number of loci 
by introducing a parameter to modify the given degrees of freedom 
and thereby informally model linkage between loci \citep[e.g.][]{EEMS}.

We estimate the values of the parameters of the model using a Bayesian approach.
Acknowledging the dependence of the parametric covariance matrix $\Omega$ on its constituent parameters
$w,\alpha,\phi,\eta,\gamma$ and on the (observed) geographic distances $D$ 
with the notation $\Omega(w,\alpha,\phi,\eta,\gamma,D)$,
we denote the posterior probability density of the parameters as:
\begin{equation}
P\left( w,\alpha,\phi,\eta,\gamma \;	| \; \widehat{\Omega}, L \right) \propto
P\left(\widehat{\Omega} \; | \; \Omega(w,\alpha,\phi,\eta,\gamma,D) \right)
P(w)P(\alpha)P(\phi)P(\eta)P(\gamma) ,
\end{equation}
where $P(w)$, $P(\alpha)$, $P(\phi)$, $P(\eta)$, and $P(\gamma)$, are prior distributions.
All parameters are given (half-)Gaussian priors 
except for $\alpha_2$, which is uniform on $(0,2)$, 
and $w$, for which we use an independent Dirichlet of dimension $K$ for each individual
(see Table \ref{tab:param_prior_tab} for specifics).
Parameters are independent between layers.
We use Hamiltonian Monte Carlo as implemented in STAN \citep{stan, NUTS, stan_lib, rstan} 
to estimate the posterior distribution on the parameters.
We also provide an R package \citep{R} called \texttt{conStruct} 
(for ``continuous structure") that functions as a wrapper around this inference machinery.

@@

A nice feature of our approach is that the model described in Eqn \eqref{cross_layer_covariance} 
contains a nonspatial assignment model as a special case 
(see \secref{model_app} for a more in-depth discussion). 
By setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is strongly conceptually similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g.][]{ADMIXTURE}; 
the only difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomially distributed \citep{Engelhardt2012}.  
Therefore, we can compare the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework. 
We also validate our claim that the results of our non-spatial method 
are a close match to those of STRUCTURE-based approaches.

@@

We use cross-validation \citep[similar in spirit to][]{ADMIXTURE_xval} to attack this problem.
To do this,
we use a ``training" partition of the data (in practice, a random 90\% subset of the loci)
to estimate the posterior distribution of the parameters,
and then calculate the log-likelihood of the remaining ``testing'' loci,
averaged over the posterior.
Prediction accuracy of a particular model (e.g., value of $K$)
is then measured using this log-likelihood,
averaged over a number of independent data partitions.
The best model is judged to be the simplest one with significantly better predictive accuracy
than others (see \secref{Xvalidation} for more on our cross-validation procedure).
In general, larger values of $K$ allows the model more flexibility,
and thus increases the likelihood of the training partition, 
but this improvement in the likelihood will plateau (or even peak), 
since above a certain $K$ the model only fits noise specific to the training data 
rather than generalizable patterns.
At any value of $K$, support for the spatial model over the nonspatial model 
means that isolation by distance is likely a feature of the data.

@@

An additional heuristic to evaluate the inclusion of additional layers is to calculate 
each layer's relative contribution to total covariance, 
and to choose a value of $K$ where all layers have a contribution above some cutoff (e.g. 0.1\%).
Because the prior on admixture proportions is a Dirichlet distribution 
that is quite harsh against intermediate admixture values (see Table \ref{tab:param_prior_tab}),  
the model will tend to ``not use" unnecessary layers, 
and they will have a low contribution to overall covariance.
The contribution of the $k$th layer can be calculated as
its contribution to the parametric covariance between samples 
summed over the upper triangle (excluding the diagonal) of the covariance matrix.
The contribution of the $k$th layer is its absolute contribution relative to those of all other layers, 
i.e., its contribution divided by the sum of the contributions of all $K$ layers
(for details on calculating this contribution, see Appendix \secref{layer_contribution}). 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

>>>>

\begin{abstract}
A classic problem in population genetics is the characterization 
of discrete population structure in the presence of 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
the use of clustering or assignment methods may incorrectly ascribe
differentiation due to continuous processes (e.g., geographic isolation by distance)
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
% This is partly a result of the difficulty of sampling uniformly and continuously 
% across the range of a population or species, but more, it 
This reflects a shortcoming of current methods for inferring and 
visualizing population structure when applied to genetic data
deriving from geographically distributed populations.
Here, we present a novel statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of two-dimensional population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This model explicitly addresses the ``clines vs. clusters" problem in 
modeling population genetic variation.
% by jointly accommodating both continuous and discrete patterns of differentiation. 
The method produces useful descriptions of structure in genetic relatedness
in situations where separated, geographically distributed populations interact,
as after a range expansion or secondary contact.
We demonstrate the utility of this approach using simulations 
and by applying it to empirical datasets of poplars and black bears in North America.
\end{abstract}

\plr{I felt like we said "continuous and discrete" too many times in the abstract.}

\newpage
\section*{Introduction}
A fundamental quandary in the description of biological diversity is
the fact that diversity shows both discrete and continuous patterns. 
For example, reasonable people can disagree about whether two 
populations are separate species, because the 
process of speciation is usually gradual, and so there is no set point
in the continuous divergence of populations when they unambiguously become discrete species.
Descriptions of taxa using phylogenetic trees omit within-taxon variation,
especially when these are intermingling subgroups.
The issue of identifying meaningful biological subunits 
extends below the species level, as patterns of phenotypic
and genetic diversity within and among populations are shaped
by continuous migration and drift, as well as by more discrete events, 
such as geographic barriers, rapid expansions, bottlenecks, and rare
long-distance migration. 
Both discrete and continuous components
are required to accurately describe most species' patterns of genetic relatedness.

From a practical standpoint, even while acknowledging continuous
processes, we often need to identify somewhat separable populations 
from which individuals are sampled \citep{wright1949genetical}.
Delineating populations is useful for systematics and for
informing conservation priorities \citep{Moritz1994,Waples_1998,Moritz_etal_2002}.
Furthermore, we often need to identify subsets of indivduals resulting from reasonably coherent
evolutionary histories for downstream analyses to learn about population history and adaptation,
\plr{as for example XXX?}
Conversely,
the substantial information available from continuous, geographic differentiation
(e.g., adaptation along a climatic gradient)
can be confounded by discrete historical processes (e.g., admixture),
requiring methods that can disentangle the two.

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \citep{CavalliSforza1975, treemix},
dimensionality-reduction approaches \citep{meirmans2009genodive}, 
such as principal components analysis 
\citep{menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches 
\citep[e.g.][]{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, 
FINESTRUCTURE, fastStructure, huelsenbeck2007inference, 
Corander2003,TESS,geneland}, among others.
Each of these methods perform best in particular situations,
but many can give misleading results when applied to data 
that show a continuous pattern of differentiation,
as that produced by geographic isolation by distance 
\citep{Wright1943, novembre_interpreting_2008, Frantz2009}.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Existing model-based clustering methods model each individual's genotypes 
as random draws from a set of underlying, unobserved population clusters, 
each with a characteristic set of allele frequencies, which are estimated. 
These underlying frequencies are identical 
for all individuals assigned to a cluster, 
\emph{regardless} of their spatial location. 
Spatial information has been incorporated into some of these methods, 
by, for example, placing spatial priors on cluster memberships \citep{geneland,TESS}, 
but this does not address the underlying issue that 
these methods assume that allele frequencies 
are constant in a cluster across the species' range.

\emph{Isolation by distance} refers to a pattern of increasing genetic differentiation
with geographic separation,
which occurs when geographically restricted dispersal allows
genetic drift to build up differentiation between distant locations
\citep{Wright1943}. 
Theoretical work,
mostly derived from ``stepping-stone'' models 
\citep{kimura1964stepping,sawyer1976stepping,shiga1984stepping},
gives us some analytical predictions for isolation by distance
\citep{malecot1969mathematics,Slatkin1985,epperson2003geographical}, 
but substantial work remains to be done \citep{barton2002neutral,barton2013modelling}.
% IBD has been modeled in both spatially continuous frameworks 
% in which an individual's parents are found within some radius of that individual's location,
% and discrete, ``stepping-stone'' frameworks \citep{kimura1964stepping},
% in which migration occurs only between neighboring demes.
Given the generality of the circumstances that generate a pattern of isolation by distance, 
it is unsurprising that IBD is very widespread in nature \citep{meirmans2012,Sexton_etal_2014}.

@@

To address this, we set out to develop
a model-based clustering method that, when possible, uses isolation by distance 
to explain observed genetic variation.
With an explicit spatial component, discrete population structure need only be invoked when genetic differentiation 
in the data deviates significantly from that expected given geographic separation.
In this paper, 
% we present a statistical method for simultaneously 
% modeling continuous and discrete patterns of population structure.
we model genetic variation in genotyped individuals as 
partitioned within or admixed across a specified number of discrete layers,
within each of which relatedness decays as a parametric function of the distance between samples.
We also implement a cross-validation approach for comparing and selecting models across different numbers of layers,
and we demonstrate the utility of our approach using both simulated and empirical data.
The implementation of this method, \texttt{conStruct} (for ``\emph{con}tinuous \emph{struct}ure"), 
is available as an R package at 
\href{https://github.com/gbradburd/conStruct}{https://github.com/gbradburd/conStruct}.

@@

\paragraph{Data}
The statistical framework of our approach is conceptually similar to that in \cite{Wasser2004}, \cite{BEDASSLE}, and \cite{spacemix},
although we use a somewhat different summary statistic than in this previous work.
The model works with allele frequencies at $L$ unlinked, bi-allelic single nucleotide polymorphisms (SNPs) genotyped across $N$ samples.
Each ``sample'' may be a single individual,  
a collection of individuals from a location, 
or frequencies estimated from pooled sequencing.
We write $N_i$ for the number of chromosomes sequenced in the $i^\text{th}$ sample.
The \emph{sample frequency} at locus $\ell$ in sample $n$, denoted $f_{n,\ell}$, 
is calculated by first arbitrarily choosing one of the observed alleles at locus $\ell$ to count, 
then dividing the number of observations of that counted allele by the total number of chromosomes genotyped at that locus
in sample $n$.
The choice of allele does not affect subsequent calculations, and so may be arbitrary.
We then calculate the \emph{allelic covariance} between samples $i$ and $j$, denoted $\widehat{\Omega}_{i,j}$,
as the expected covariance of distinct individual alleles 
chosen from each of the two samples at a random locus,
coding the counted allele as `1' (and the other as `0').
Concretely, this is
\begin{equation}
\widehat{\Omega}_{i,j} = 
    \frac{1}{L} \sum_{\ell=1}^L \left( (f_{i,\ell}-1/2) (f_{j,\ell}-1/2) - \delta_{i,j} f_{i,\ell}^2 / (N_i-1) \right),
\label{allelic_covariance}
\end{equation}
where $\delta_{i,j}=1$ if $i=j$ and is 0 otherwise.
Although we describe this as a covariance between individually drawn alleles,
$\widehat{\Omega}_{i,j}$ is in fact also the covariance between the allele frequencies
of a randomly chosen allele in samples $i$ and $j$, as long as $i \neq j$.
The diagonal (where $i=j$) does not have this interpretation, 
and reflects the covariance between individual alleles drawn from within the population.

@@

\paragraph{Continuous and discrete differentiation}
Clustering approaches to describing genetic variation
are useful because population history can often be meaningfully described on a coarse scale
by interactions between discrete ``populations''
whose relationships are delimited by patterns of glaciation,
large-scale migration,
mountain ranges, and the like.
Here we add a spatial component within each such discrete historical component,
which we refer to as a set of ``layers'' that overlay the modern map.
We imagine each layer as a geographically distributed population 
that extends over the entire sampled range of the populations. 
As depicted in Figure \ref{schematic},
each sample is composed of a mixture of contributions from each of these layers,
with the relative contributions of each layer described by
a set of ``admixture proportions'' (the $w_i^{(k)}$).
These layers thus take the place of ``clusters'' in clustering methods,
but we do not adopt this term, 
as ``spatial cluster'' suggests a clustering in space, 
while our layers may contribute to genetic variation 
across the entire geographic range.

@@

Within each of these layers,
allele frequencies have positive covariance at geographically close locations,
but this covariance is allowed to decay as geographic distance increases.
This pattern of spatial decay reflects how migration between neighboring demes  
homogenizes allele frequency changes that arise locally due to drift, 
but less effectively homogenizes geographic distant demes,
resulting in a continuous pattern of isolation by distance within each layer.
There is a fixed amount of covariance between layers, irrespective of spatial location.
Within each layer allele frequencies are expected to change gradually with distance,
but observed frequencies can change abruptly at many loci 
if the proportions by which individuals' ancestry derive from each layer 
(the ``admixture'' proportions) 
do so as well.

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each layer, 
we define the covariance within layer $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_layer_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th layer.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
and the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the cluster.
Our choice of a powered-exponential decay, 
as parameterized by the $\alpha$s, 
is a flexible and standard choice in spatial statistics \citep{Diggle1998}, 
and is not chosen to match a particular population genetics model. 
% (Analytical theory \citep{barton2002neutral} predicts the decay should be proportional to $\exp{-D}/\sqrt{D}$.)
The $\phi^{(k)}$ is a parameter that describes the background covariance within the cluster. 
If two samples draw 100\% of their ancestry from cluster $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the cluster.
%% plr: this is no longer true
% The ``shared drift'' parameter $\phi^{(k)}$ can be interpreted as the branch length 
% connecting the $k$th population to the population ancestral to all modeled
% layers \citep[see for example ][]{patterson_ancient_2012,peter_fstats}. 
In the methods section \ref{rationale} we lay out a
simple model of allele frequencies underlying this covariance model.

We then allow samples to draw their ancestry from more than one layer.
The ``admixture'' proportion of the $i$th sample in the $k$th layer, denoted $w^{(k)}_i$,
gives the genome-wide proportion of alleles from sample $i$ that derive from
layer 
$k$ (and so $\sum_{k=1}^K w^{(k)}_i =1$).
A visual representation of the method mechanics is shown in Fig \ref{schematic}.

\begin{figure}
	\centering
		{\includegraphics[width=4in,height=2.5in]{figs/schematic/method_schematic.png}}
		\caption{Schematic of our method, using $K=3$ as an example.
			     The pattern of isolation by distance within each layer is denoted by their color gradients, 
			     and $\phi^{(k)}$ denotes the covariance shared by samples with ancestry entirely in the $k$th layer.
			     Sampled populations on the landscape are inferred to be admixed between these layers; 
			     the $i$th sample draws proportion $w^{(k)}_i$ of its ancestry from layer $k$.
			     For convenience, each layer is depicted as a small square, but in fact, 
			     each layer exists everywhere in the sampled area,
			     so the small dashed circles on each layer show where 
			     the location of the highlighted, admixed sample intersects each layer.
			    }\label{schematic}
\end{figure}

We can then describe the covariance between samples $i$ and $j$ across all layers, $\Omega_{i,j}$,
by summing their within-layer spatial covariances ($G_{i,j}^{(k)}$ in layer $k$) across all $K$ layers,
weighted by the relevant admixture proportions.
\begin{equation}
    \Omega_{i,j} = \gamma + \sum_{k=1}^K w^{(k)}_i w^{(k)}_j
G^{(k)}_{i,j} + \delta_{i,j}\eta_i .
\label{cross_layer_covariance}
\end{equation}
In this equation, $w^{(k)}_i w^{(k)}_j$ is the proportion of alleles that \emph{both}
sample $i$ and sample $j$ have inherited from layer $k$.

@@

\paragraph{Likelihood and inference}
If the allele frequencies at each locus
were independent between loci and multivariate normally distributed across populations, 
their allelic covariance $\widehat{\Omega}$ 
would be Wishart distributed with degrees of freedom equal to $L$, 
the number of loci genotyped.
We use this as a convenient approximation to the true distribution described above,
and so define the likelihood of the allelic covariance to be
\begin{equation}
    P(\widehat{\Omega} \; | \; \Omega) 
    = 
    \mathcal{W} \left( L\widehat{\Omega} \; | \; \Omega,L\right) ,
\end{equation}
where $\mathcal{W}$ is the Wishart likelihood function.
Statistical nonindependence between loci (linkage disequilibrium)
will decrease the effective
number of degrees of freedom.  
One possible solution, which we have not yet found necessary to implement,
would be to estimate an \emph{effective} number of loci 
by introducing a parameter to modify the given degrees of freedom 
and thereby informally model linkage between loci \citep[e.g.][]{EEMS}.

We estimate the values of the parameters of the model using a Bayesian approach.
Acknowledging the dependence of the parametric covariance matrix $\Omega$ on its constituent parameters
$w,\alpha,\phi,\eta,\gamma$ and on the (observed) geographic distances $D$ 
with the notation $\Omega(w,\alpha,\phi,\eta,\gamma,D)$,
we denote the posterior probability density of the parameters as:
\begin{equation}
P\left( w,\alpha,\phi,\eta,\gamma \;|\; \widehat{\Omega} \right) 
    \propto
    P\left(\widehat{\Omega} \; | \; \Omega(w,\alpha,\phi,\eta,\gamma,D) \right)
    P(w)P(\alpha)P(\phi)P(\eta)P(\gamma) ,
\end{equation}
where $P(w)$, $P(\alpha)$, $P(\phi)$, $P(\eta)$, and $P(\gamma)$, are prior distributions.
All parameters are given (half-)Gaussian priors 
except for $\alpha_2$, which is uniform on $(0,2)$, 
and $w$, for which we use an independent Dirichlet of dimension $K$ for each individual
(see Table \ref{tab:param_prior_tab} for specifics).
Parameters are independent between layers.
We use Hamiltonian Monte Carlo as implemented in STAN \citep{stan, NUTS, stan_lib, rstan} 
to estimate the posterior distribution on the parameters.
We also provide an R package \citep{R} called \texttt{conStruct} 
(for ``continuous structure") that functions as a wrapper around this inference machinery.

@@

A nice feature of our approach is that the model described in Eqn \eqref{cross_layer_covariance} 
contains a nonspatial assignment model as a special case 
(see \secref{model_app} in the appendix for a more in-depth discussion). 
By setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is very similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g.][]{ADMIXTURE}; 
the only difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomially distributed \citep{Engelhardt2012}.  
(We make this approximation for the substantial advantages in computational speed.)
Therefore, we can compare the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework. 
We also validate our claim that the results of our non-spatial method 
are a close match to those of STRUCTURE-based approaches.

@@

We use cross-validation \citep[similar in spirit to][]{ADMIXTURE_xval} to attack this problem.
To do this,
we use a ``training" partition of the data (in practice, a random 90\% subset of the loci)
to estimate the posterior distribution of the parameters,
and then calculate the log-likelihood of the remaining ``testing'' loci,
averaged over the posterior.
Prediction accuracy of a particular model (e.g., value of $K$)
is then measured using this log-likelihood,
averaged over a number of independent data partitions.
The best model is judged to be the simplest one with significantly better predictive accuracy
than others (see \secref{Xvalidation} of the appendix for more on our cross-validation procedure).
In general, larger values of $K$ allows the model more flexibility,
and thus increases the likelihood of the training partition, 
but this improvement in the likelihood will plateau (or even peak), 
since above a certain $K$ the model only fits noise specific to the training data 
rather than generalizable patterns.
At any value of $K$, support for the spatial model over the nonspatial model 
means that isolation by distance is likely a feature of the data.

@@

Another way to describe how additional layers contribute to model fit
is to calculate each layer's relative contribution to total covariance, 
and to choose a value of $K$ where all layers have a contribution above some cutoff (e.g. 0.1\%).
The Dirichlet prior on admixture proportions
is quite harsh against intermediate admixture values (see Table \ref{tab:param_prior_tab}),  
encouraging the model to ``not use" unnecessary layers, 
so that they will have a low contribution to overall covariance.
We use the connection of allelic covariance to the probability of agreement (identity) between genotypes
to calculate a measure of overall contribution of each layer to the model.
XXX PLR XXX
To calculate the overall contribution of the $k$th layer to the model,
we use the connection between allelic covariance and

The contribution of the $k$th layer can be calculated as
its contribution to the parametric covariance between samples 
summed over the upper triangle (excluding the diagonal) of the covariance matrix.
The contribution of the $k$th layer is its absolute contribution relative to those of all other layers, 
i.e., its contribution divided by the sum of the contributions of all $K$ layers
(for details on calculating this contribution, see Appendix \secref{layer_contribution}). 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.
